---
title: States, War, and Conflict Patterns
subtitle: POSC 3610  -- International Conflict
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: false
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}

library(tidyverse)
library(stevemisc)
library(countrycode)
library(knitr)
# library(maddison)

InterWars <- read_csv("~/Dropbox/data/cow/wars/Inter-StateWarData_v4.0.csv") 
ExtraWars <- read_csv("~/Dropbox/data/cow/wars/Extra-StateWarData_v4.0.csv") 
IntraWars <- read_csv("~/Dropbox/data/cow/wars/Intra-StateWarData_v4.1.csv") 

MIDA <- read.csv("~/Dropbox/projects/mid-project/gml-mid-data/2.03/gml-mida-2.03.csv")
MIDB <- read.csv("~/Dropbox/projects/mid-project/gml-mid-data/2.03/gml-midb-2.03.csv")

DDY <- read_csv("~/Dropbox/projects/mid-project/gml-mid-data/2.03/gml-ddy-disputes-2.03.csv")
States <- read_csv("~/Dropbox/data/cow/states/states2016.csv")

NMC <- read_csv("~/Dropbox/data/cow/cinc/NMC_5_0.csv")

States %>%
  mutate(endyear = ifelse(endyear == 2011, 2015, endyear)) %>%
  rowwise() %>%
  mutate(year = list(seq(styear, endyear))) %>%
  ungroup() %>%
  unnest() %>%
  arrange(ccode, year) %>%
  select(ccode, year) %>%
  distinct(ccode, year) -> CYs


```

# Introduction

<!-- ### What's the Diehl (2006) article about?

Someone tell me what this article is about...

- Btw, I will do this every class lecture. Get used to it. 

### Introduction

This is actually a presidential address to Peace Science Society.

- It summarizes the state of conflict research for the 2005 annual meeting.

For the purpose of this class, Diehl asks "what is inter-state conflict?"

- He provides a nice survey of the conflict literature too. -->

# Sarkees et al. (2003)
## Whither War?
### Whither War?

Motivating question: what is the future of war after the Cold War?

- May seem quaint, but it was an interesting question at the time.

Is war on the verge of obsolence?

- Period after WWII rather calm, all things considered.
- Cold War only time when two largest countries did not fight each other.
	- True going back to the ancient Greeks.
- Also features concurrent emergence of "the democratic peace."
- Numerous territorial claims settled through 1980s/90s (e.g. Ecuador-Peru).

Did the end of the Cold War mark the "end of history?"

### Whither War?

We might have evidence that war is actually increasing.

- End of Cold War: emergence of "ethnic wars" (e.g. Rwanda, Yugoslavia).
- There might be a new "clash of civilizations."

### What Do We Make of These?

Is war over? Is it increasing?

- Does it depend on how we define war and what data we use?

Maybe, and this motivates the Sarkees et al. (2003) article.

- Defining war is important, and we will do it here.

### Making Our Terms Clear

A lot of people are going to define war differently.

- Do massacres count as "war?"
- Was the Cold War a "war?" How about the "Cod (sic) War?"
- Are the Koreas still at war? The North says it is.

We need a hard definition.

- We also need a *classification*.

## What is a State?
### What is a "State?"

We need to start with a "state" first.

- In lay terms, we call it a "country."

This actually kind of begs the question though. What is it?

- Is the Vatican a "state?"
- What about Vichy France?

### Defining a "State"

Definitions changed throughout the history of the project, but our conceptualization of a state must satisfy five criteria: 

1. A state is a "geopolitical entity." It must occupy territory.
	- For example, the U.N. is not a state by this criterion.
	- South Vietnam after the fall of Hanoi is not either.
2. A state must have a population above 500,000.
	- Entities like Lichtenstein become "microstates" by this criterion.
3. A state must be diplomatically recognized writ large.
	- Before 1919: this was recognition by the U.K. or France.
	- After 1919: League of Nations or, now, the U.N.

### Defining a "State"

4. A state must be independent in its own affairs.

	- This criterion sorts out entities like Puerto Rico and Hong Kong.
	- These are subsidiaries of other states.
	- Importantly, they don't have their own foreign policies.

5. A state must have a sovereign political authority.

	- It doesn't have to be democratically elected, but one must exist.
	- This criterion tries to weed out failed states and puppet states.

### Examples of States and Non-states

Here are some examples of a state and non-states by this classification.

- Examples come from Sarkees and Wayman (2010).

\scriptsize

| Entity | Territory | Population | Recognition | Sovereign | Independence | State? |
|:-------|:---------:|:----------:|:-----------:|:---------:|:------------:|:------:|
| USA    | Yes | Yes | Yes | Yes | Yes | **Yes** |
| Vatican | Yes | No | Yes | Yes | Yes | **No**  |
| Confederacy | Yes | Yes | No | Yes | Yes | **No** |
| Vichy France | Yes | Yes | Yes | No | No | **No** |
| Ukrainian SSR | Yes | Yes | Yes | Yes | No | **No** |

\normalsize

## What is a War?
### What is a "War?"

We know what a state is, now let's classify wars.

1. Inter-state war
	- These are wars between (at least) two states.
	- Examples: Franco-Prussian War, World War II
2. Intra-state war
	- These are between a state government and domestic group.
	- These are typically fought for control of government or secession.
	- Examples: American Civil War, Angolan Civil War
3. Extra-state war
	- These are wars between a state government and an independent non-state actor or dependent non-state actor.
	- Typically known as "imperial war" or "colonial war," respectively.
	- Examples: Opium War (1839-1842), French-Indochina War (1946-1954).

### The Threshold

Importantly, we define wars as a *fatality threshold.*

- Any conflict that arises in seriousness to 1,000 battle-related deaths per year is a war.
- We take care to delineate *battle*-deaths from massacres.
	- In other words: we only count troops.
	- via old colleagues at Alabama: you'd be surprised how rare civilian fatalities are.
	
### The Threshold

Disputes can conceptually meet this threshold and then drop back to lower levels of seriousness.

- The dispute would still continue.
- In practice, this is rare in the inter-state context. More common in the intra-state context.
- It wouldn't affect the MID data, but it'd affect the war data.

# Patterns in MIDs and War
###

```{r cowinterwarsdecade, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

InterWars %>% distinct(WarName, .keep_all=TRUE) %>%
  mutate(decade = paste0(floor(StartYear1/10)*10,"s")) %>%
  group_by(decade) %>% summarise(numwars = n()) %>%
  ggplot(., aes(decade, numwars)) + theme_steve_web() +
  geom_bar(stat="identity", color="black", alpha=0.8, fill="#619cff") +
  xlab("Decade") + ylab("Number of Inter-State Wars") +
  geom_text(aes(label=numwars), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  scale_y_continuous(breaks=c(0, 2, 4, 6, 8, 10)) +
  labs(caption="Data: Correlates of War (Inter-State) War Data (v. 4.0). Temporal Domain: 1816-2007.",
       title = "Number of Inter-State War Onsets, by Decade",
       subtitle = "The 1910s saw the most inter-state war onsets of any decade in the data, followed by nine onsets in the 1970s.")

```

###

```{r gmlmidwarspercyear, eval=T, echo=F, fig.height=8.5, fig.width=14, message = F}

CYs %>%
  group_by(year) %>%
  summarize(numstates = n()) -> Numstates

DDY %>%
  filter(hostlev == 5) %>%
  distinct(ccode1, year, .keep_all=TRUE) %>%
  group_by(year) %>%
  summarize(numstateswars = n()) %>%
  left_join(Numstates, .) %>%
  mutate(numstateswars = ifelse(is.na(numstateswars), 0, numstateswars),
         perc = mround2(numstateswars/numstates)) %>% 
  ggplot(.,aes(year, perc)) + theme_steve_web() + 
  geom_bar(stat="identity", fill="#f8766d", alpha=0.8, color="black") +
  xlab("Year") + ylab("Percentage of the State System") +
  scale_x_continuous(breaks = seq(1820, 2010, by = 10)) +
  labs(title = "The Percentage of States Involved in Interstate War by Year, 1816-2010",
       subtitle = "We treat interstate war as (fortunately) a rare event but the 1860s, 1910s, and 1940s stand out as particularly violent decades.",
       caption = "Data: GML MID data (v. 2.02) and Correlates of War State System Membership List.")



```

###

```{r cowinterwarsdeaths, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

InterWars %>%
  mutate(BatDeath = ifelse(BatDeath == -9, 1000, BatDeath)) %>%
  mutate(decade = paste0(floor(StartYear1/10)*10,"s")) %>%
  group_by(decade) %>% summarise(batdeath = sum(BatDeath)) %>%
  ggplot(., aes(decade, batdeath)) + theme_steve_web() +
  geom_text(aes(label=prettyNum(batdeath, big.mark=",", scientific=F)), vjust=-.5, colour="black",
            position=position_dodge(.9), size=3.5, family="Open Sans") +
  geom_bar(stat="identity", color="black", fill="#f8766d", alpha = 0.8) +
  xlab("Decade") + ylab("Number of Battle Deaths") +
  labs(caption="Data: Correlates of War (Inter-State) War Data (v. 4.0). Temporal Domain: 1816-2007.",
       subtitle = "Most wars aren't that violent, per se. World War I and World War II? Those were *really* violent.",
       title = "Number of Inter-State War Battle Deaths, by Decade")

```


<!-- ###

```{r cowinterwarsdeathsperc, echo=F, eval=F, fig.width = 14, fig.height = 8.5}

NMC %>%
  mutate(decade = paste0(floor(year/10)*10,"s")) %>%
  group_by(decade, ccode) %>%
  summarize(cavgpop = mean(tpop, na.rm=T)) %>%
  group_by(decade) %>%
  summarize(dtpop = sum(cavgpop, na.rm=T)) -> decadepop

InterWars %>%
  mutate(BatDeath = ifelse(BatDeath == -9, 1000, BatDeath)) %>%
  mutate(decade = paste0(floor(StartYear1/10)*10,"s")) %>%
  group_by(decade) %>% summarise(batdeath = sum(BatDeath)) %>%
  mutate(batdeath = batdeath/1000) %>%
  left_join(., decadepop) %>%
  mutate(perc = batdeath/dtpop) %>%
  ggplot(., aes(decade, perc)) + theme_steve_web() +
  geom_text(aes(label=prettyNum(batdeath, big.mark=",", scientific=F)), vjust=-.5, colour="black",
            position=position_dodge(.9), size=3.5, family="Open Sans") +
  geom_bar(stat="identity", color="black", fill="#f8766d", alpha = 0.8) +
  xlab("Decade") + ylab("Number of Battle Deaths") +
  labs(caption="Data: Correlates of War (Inter-State) War Data (v. 4.0). Temporal Domain: 1816-2007.",
       subtitle = "Most wars aren't that violent, per se. World War I and World War II? Those were *really* violent.",
       title = "Number of Inter-State War Battle Deaths, by Decade")

```

-->

###

```{r cowextrawarsdecade, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

ExtraWars %>% distinct(WarName, .keep_all=TRUE) %>%
  mutate(decade = paste0(floor(StartYear1/10)*10,"s")) %>%
  group_by(decade) %>% summarise(numwars = n()) %>%
  ggplot(., aes(decade, numwars)) + theme_steve_web() + 
  geom_bar(stat="identity", fill="#f8766d", alpha=0.8, color="black") +
  geom_text(aes(label=numwars), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  xlab("Decade") + ylab("Number of Extra-State Wars") +
  scale_y_continuous(breaks=c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22)) +
  labs(caption="Data: Correlates of War (Extra-State) War Data (v. 4.0). Temporal Domain: 1816-2007.",
       title = "Number of Extra-State War Onsets, by Decade",
       subtitle = "Extra-state wars are mostly historical phenomenon in the Correlates of War data, disappearing as the state system expanded after the end of empires.") +
  geom_rect(xmin=14.5, xmax=20.1, ymin=18.75, ymax=Inf, fill="white", color="black", linetype="dotted") +
  annotate("text", x=20, y=21, 
           label="1980s: The Soviet-Afghanistan Quagmire (1980-1989)\n1990s: Iraqi-Kurdish Conflict (Turkey: 1991-92, 1997)\n2000s: Al Aqsa Intifada (2000-3)\n     Afghan Resistance (2001-forever)\nIraqi Resistance (2003-probably forever as well)", size=3.5, family="Open Sans", hjust=1, fontface = "bold")

```

###

```{r cowintrawarsdecade, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

IntraWars %>% distinct(WarName, .keep_all=TRUE) %>%
  mutate(decade = paste0(floor(StartYear1/10)*10,"s")) %>%
  group_by(decade) %>% summarise(numwars = n()) %>%
  ggplot(., aes(decade, numwars)) + theme_steve_web() +
  geom_text(aes(label=numwars), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  geom_bar(stat="identity", alpha=0.8, fill="#f8766d", color="black") +
  xlab("Decade") + ylab("Number of Intra-State Wars") +
  scale_y_continuous(breaks=seq(0, 40, 4)) +
  labs(caption="Data: Correlates of War (Intra-State) War Data (v. 4.1). Temporal Domain: 1816-2007.",
       title = "Number of Intra-State War Onsets, by Decade",
       subtitle = "Most of the empirical evidence against the 'End of History' arguments focused on the explosion of civil wars after the Cold War.")

```

###

```{r gmlmidhostlev, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

MIDA %>%  group_by(hostlev) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  mutate(hostlev = c("2: Threat to\nUse Force", "3: Display of Force", "4: Use of Force", "5: War")) %>%
  ggplot(., aes(hostlev, count)) + theme_steve_web() +
  geom_bar(stat="identity", fill="#619cff", color="black", alpha=0.8)  +
  geom_text(aes(label=count), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  ylab("MID Count") + xlab("Hostility Levels") +
  labs(caption="Data: Gibler-Miller-Little MID Data (v 2.03), 1816-2010.",
       title="Number of MIDs by Dispute Hostility Level",
       subtitle = "Most MIDs involve at least some of use of force, which constitutes the highest action of 66.89% of the data set.")
  

```

###
```{r gmlmidhiact, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

MIDA %>%  group_by(hiact) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(., aes(as.factor(hiact), count)) + theme_steve_web() +
  geom_bar(stat="identity", alpha=0.8, color="black", fill="#619cff")  +
  geom_text(aes(label=count), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  scale_x_discrete(labels = c("Threat to\nUse Force", "Threat to\nBlockade",
                              "Threat to\nOccupy\nTerritory",
                              "Threat to\nDeclare War",
                              "Show of\nForce", "Alert",
                              "Mobilization", "Border\nFortification",
                              "Border\nViolation","Blockade",
                              "Occupation\nof\nTerritory", "Seizure",
                              "Attack","Clash", "Declaration\nof\nwar",
                              "Use of\nCBRN\nWeapons", "Begin\nInterstate\nWar",
                              "Join\nInterstate\nWar")) +
  geom_vline(xintercept=4.5, linetype="dashed") +
  geom_vline(xintercept = 9.5, linetype="dashed") +
  geom_vline(xintercept = 16.5, linetype="dashed") +
  annotate("text", x= 16.25, y=590, family="Open Sans", hjust=1, size=3.5, label="Use of Force", fontface="bold") +
  annotate("text", x= 9.25, y=590, family="Open Sans", hjust=1, size=3.5, label="Show of Force", fontface="bold") +
  annotate("text", x= 4.25, y=590, family="Open Sans", hjust=1, size=3.5, label="Threat of Force", fontface="bold") +
  annotate("text", x= 18.25, y=590, family="Open Sans", hjust=1, size=3.5, label="War", fontface="bold") +
  ylab("MID Count") + xlab("Highest Level of Action") +
  labs(caption="Data: Gibler-Miller-Little MID Data (v. 2.03), 1816-2010.
       Note: MID#1109 is unique for being the only dispute in the data for which the highest action was the use of CBRN weapons. Other CBRN weapon uses (e.g. World War II, Iran-Iraq War) happen during war itself, a higher hostility level.",
       title="Number of MIDs by Highest Action Observed at Dispute-Level",
       subtitle = "Most common hiacts: Attack (n: 577), Clash (n: 480), Show of Force (n: 307), Seizure (n: 261)")
  

```

###
```{r gmlmidoutcomes, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

MIDA %>%  group_by(outcome) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  mutate(outcome = c("Missing/\nUnknown", "Victory for\nSide A", "Victory for\nSide B",
                     "Yield by\nSide A", "Yield by\nSide B", "Stalemate",
                     "Compromise", "Released", "Unclear", "Joins\nOngoing War")) %>%
  ggplot(., aes(as.factor(outcome), count)) + theme_steve_web() +
  geom_bar(stat="identity", alpha=0.8, color="black", fill="#619cff")  +
  geom_text(aes(label=count), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  ylab("MID Count") + xlab("Different Outcomes") +
  labs(caption="Data: Gibler-Miller-Little MID Data (v 2.03), 1816-2010.",
       title="How MIDs End",
       subtitle = "By far, most MIDs---around 63%---end in a stalemate.")
  

```

###

```{r gmlmidlength, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

MIDA %>%
   mutate(category=cut(mindur, breaks=c(-Inf, 1, 30,  120, 365, 730, 1095, Inf), 
                       labels=c("One Day","2-30\nDays\n(i.e. a month)", 
                                "31-120\nDays\n(i.e. 1-4 months)",
                                "121-365\nDays\n(i.e. 4-12 months)",
                                "1-2 Years", "2-3 Years", "Longer Than\n3 Years"))) %>%
  group_by(category) %>% summarize(count = n()) %>%
  ggplot(.,aes(category, count)) + 
  geom_bar(stat="identity", alpha=0.8, color="black", fill="#619cff")  +
  theme_steve_web() +
  geom_text(aes(label=count), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  xlab("Minimum Durations (condensed)") + ylab("MID Count") +
  labs(caption="Data: Gibler-Miller-Little MID Data (v. 2.03), 1816-2010.",
       title="Most MIDs Don't Last That Long",
       subtitle = "Over 28% of MIDs last a single day. More than half of all MIDs last 30 days or fewer.")
  

```


###

```{r gmlmidlongestmids, echo=F, eval=T}
MIDA %>%
  arrange(-mindur) %>%
  select(dispnum3, styear, endyear, mindur) %>% head(11) %>%
  mutate(Description = c("Israel v. Lebanon/Syria (re: Hezbollah)",
                         "Intervention in Angolan Civil War", 
                         "Intervention in Laotian Civil War", 
                         "Vietnam War", 
                         "Cambodian–Vietnamese War",
                         "Korean DMZ Conflict",
                         "Armenia v. Azerbaijan",
                         "Iran-Iraq War",
                         "Israel v. Jordan (re: Palestine)",
                         "Paraguayan War",
                         "I'll let you guess this one...")) %>%
  rename(`MID #` = dispnum3, `Start Year` = styear,
         `End Year` = endyear, Duration = mindur) %>%
  kable(., align=c("l","c","c","c","l"),
        caption = "The Longest MIDs in History (Data: Gibler-Miller-Little MID Data (v. 2.03), 1816-2010)")
```


###

```{r gmlmidmostmids, echo=F, eval=T}
MIDB %>%
  group_by(ccode) %>%
  summarize(sum = n()) %>%
  arrange(-sum) %>% head(10) %>%
  mutate(ccode = countrycode(ccode,"cown","country.name"),
         ccode = ifelse(ccode == "United Kingdom of Great Britain and Northern Ireland", "United Kingdom", ccode),
         ccode = ifelse(ccode == "Russian Federation", "Russia/USSR", ccode),
         ccode = ifelse(ccode == "Iran (Islamic Republic of)", "Iran", ccode)) %>%
  rename(Country = ccode, `Number of MIDs All-Time` = sum) %>%
  kable(., align=c("l", "c"),
        caption = "All-Time Participation in MIDs (Data: Gibler-Miller-Little MIDs, 1816-2010)")
  
```

###

```{r gmlusamidsdecade, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

MIDB %>%
  filter(ccode == 2 | ccode == 365) %>%
  mutate(Country = ifelse(ccode == 2, "United States", "Russia/USSR")) %>%
  mutate(decade = paste0(floor(styear/10)*10,"s")) %>%
  group_by(decade, Country) %>% summarise(nummids = n()) %>%
  bind_rows(., tibble(decade = "1810s", Country = "Russia/USSR", nummids = 0)) %>%
  bind_rows(., tibble(decade = "1820s", Country = "United States", nummids = 0)) %>%
  bind_rows(., tibble(decade = "2010s", Country = "Russia/USSR", nummids = 0)) %>%
  # filter(decade != "NAs") %>%
    ggplot(., aes(decade, nummids, color=Country)) + theme_steve_web() + 
  geom_text(aes(label=nummids, group=Country), color="black",
            position=position_dodge(width=.9), size=4, vjust = -.5) +
  geom_bar(aes(fill = Country), position = "dodge", stat="identity", alpha = I(0.8),color = I("black")) +
  xlab("Decade") + ylab("Number of MIDs") +
  scale_y_continuous(breaks=seq(0, 45, 5)) +
  labs(caption="Data: Gibler-Miller-Little MID Data (v. 2.03), 1816-2010.",
       title = "U.S. and Russian MID Participation, by Decade",
       subtitle = "Russia/USSR had a more active/bellicose foreign policy in the interwar years than the U.S. while the U.S. effectively surpassed Russia in MID count in the 1980s.")

```

# Conclusion
### Conclusion

War isn't "decreasing" or "increasing."

- It's changing by type.
- Civil wars are more common now than inter-state wars.
- Extra-state wars are mostly things of the past.

MIDs have several interesting patterns as well.

- Most involve the use of force.
- Almost half become clashes or attacks.
- Most MIDs don't last long, and most end in stalemates.