---
title: "Wrapping Up: What Do We Know About International Conflict?"
subtitle: POSC 3610  -- International Conflict
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: false
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)

library(tidyverse)
library(stevemisc)
library(countrycode)
library(knitr)
# library(maddison)

Disputes <- read_csv("~/Dropbox/projects/mid-project/gml-mid-data/2.03/gml-ndy-disputes-2.03.csv")

```

# Introduction
### Puzzle for Today

*What do we know about war? And what's still left as a matter of disagreement?*

# What Do We Know About War?
## Received Wisdom and General Consensus
### Dangerous Dyads

Bremer's (1992) "dangerous dyads" still holds up well.


1. Contiguity
2. Joint democracy
3. Power preponderance
4. Major powers
5. Joint alliance
6. Advanced economies
7. Militarization

###

![](../dangerous-dyads/figs/dangerous-dyads-dwplot.pdf)

###

```{r, eval=T, echo=F, fig.height=8.5, fig.width=14, message = F, warning=F}
tribble(
  ~period, ~Type, ~perc, ~count,
  #-----, -------, -----, -----
  "1648-1714", "Territory", 77, 17,
  "1648-1714", "Territory + Territory-related", 86, 19,
  "1648-1714", "Other Issue", 14, 3,
  "1715-1814", "Territory", 72, 26,
  "1715-1814", "Territory + Territory-related", 83, 4+26,
  "1715-1814", "Other Issue", 17, 6,
  "1815-1914", "Territory", 58, 18,
  "1815-1914", "Territory + Territory-related", 84, 18+8,
  "1815-1914", "Other Issue", 16, 5,
  "1918-1941", "Territory", 73, 22,
  "1918-1941", "Territory + Territory-related", 93, 28,
  "1918-1941", "Other Issue", 7, 2,
  "1945-[1990]", "Territory", 47, 27,
  "1945-[1990]", "Territory + Territory-related", 79, 27+19,
  "1945-[1990]", "Other Issue", 21, 12
) %>%
  mutate(Type = forcats::fct_relevel(Type, "Territory", "Territory + Territory-related", "Other Issue"),
         perc = perc/100) %>%
  ggplot(.,aes(x=period, y=perc, fill=Type)) + theme_steve_web() +
  geom_bar(stat="identity", position = "dodge",
           alpha = I(0.8), color="black") +
  scale_fill_brewer(palette="Set1") +
  xlab("Historical Period") + ylab("Percentage of All Wars") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percentage and Frequency of Wars By Issue Type, 1648-1990",
       subtitle = "Most wars over time have been fought over territory or territory-related issues than other issue types.",
       caption = "Data: Vasquez (1993) via Holsti (1991). Note: counts appear on top of the bars by issue-type.") +
  geom_text(aes(label = count, group = Type), color="black",
            position = position_dodge(width=.9), size=4,
            vjust = -.5, family ="Open Sans")

```


### Democracies Are a Peculiar Class of Countries

We'll note disagreement later, but generally democracies are a unique class of countries.

- Democracies find ways to avoid war with each other.
- Democracies tend to perform well in war.
- However, democracies are no more or less war-prone than other state types.

That we even observe this defies received wisdom from our structural theories.

### War is Bargaining

We generally see war as bargaining failure (or, alternatively, bargaining via other means).

- Issue indivisibility
- Commitment problems
- Asymmetric information

### When Does Deterrence Work?

Deterrence is a long-standing debate, but we tend to believe it works the more credible threats are.

- Unilateral deterrence offers more paths to peace.

Worth noting: credible threats make deterrence more likely to succeed, but does not guarantee it.

- It won't guarantee it if challengers are highly motivated.

### Not All Wars Look Like the "Big Ones"

...and thank god, but it does mean we're mindful of what we're studying. Generally:

- Expansion follows opportunity and willingness.
- Easier to rethink it as an expected utility calculation.

### Not All Wars Look Like the "Big Ones" (Continued)


Further correlates of war's cost and outcomes:

- "Rougher" terrain
- Balance among disputants
- More troops in the combat zone.
- *Fewer* states involved

One caveat: modeling "costs" of war is tricky, and we tend to do it ex post.

### Conflict and War are Concentrated in a Handful of Countries

Wars are definitely not "independent and identically distributed." There are usual suspects.

- Rivalry explains a big part of this.
- Commitment problems following bargaining

###

```{r, eval=T, echo=F, fig.height=8.5, fig.width=14, message = F, warning=F}
Disputes %>%
  group_by(dispnum) %>%
  distinct(dispnum, .keep_all=T) %>%
  mutate(war = ifelse(hostlev == 5, 1, 0),
         fatal = ifelse(fatality1 == 1 | fatality2 == 1, 1, 0)) %>%
  group_by(ccode1, ccode2) %>%
  summarize(`No. of MIDs` = n(),
            `No. of Wars` = sum(war),
            `No. of Fatal MIDs` = sum(fatal)) %>%
  arrange(-`No. of MIDs`) %>%
  mutate(`Country A` = countrycode::countrycode(ccode1, "cown", "country.name"),
         `Country B` = countrycode::countrycode(ccode2, "cown", "country.name")) %>%
  ungroup() %>% select(-ccode1, -ccode2) %>%
  select(`Country A`, `Country B`, `No. of MIDs`:`No. of Fatal MIDs`) %>% head(10) %>%
  kable(.,
        caption = "Dyads With the Most MIDs (and Type of MIDs), 1816-2010 (GML MID Data, v. 2.03)")
  
```

## Areas of Contention
### Are Alliances War-Prone or Paths to Peace?

There is still a fair bit of debate about what role alliances play in conflict onset.

- Recent exchanges between Leeds and Johnson (2017) and Vasquez et al. (2017) highlight this disagreement.

You generally saw this in our lectures.

- No relationship between alliance proliferation and war at system-level.
- Allies are unlikely to fight each other.
- Some allies *are* likely to fight each other.

### Are Conflict Processes the Same for Onset and Escalation?

Recall Diehl (2006) encouraged us to think of different phases.

- Different phases may have different contextual influences.

We generally don't want to stray too far from a "dangerous dyads" framework, but we've found:

- Power parity leads to MIDs, but MIDs between equals are less likely to lead to war.
- Joint democracy leads to fewer MIDs, not necessarily fewer wars.
- Satisfied states have fewer MIDs, not necessarily fewer wars.
- Allies still have disputes; just unlikely to escalate to war.


### Power Matters, but How?

"Power" is a central topic in IR, but how it matters is unclear.

- Is it the contested benefit or the source of bargaining breakdown?
- Simple explanations of power via CINC do poorly.

Whatever role "power" has in conflict onset, we tend to eschew "neorealist" explanations.

### Does the Democratic Peace Have Anything to Do With Democracy?

Democratic peace might be the most important finding in IR, but doubts persist:

- "Capitalist peace" (ed. there are a laundry list of problems with this argument)
- Common systems or common interests? (ed. we don't adequately model temporal variation)
- Territorial peace (i.e. have we put the cart before the horse?)

In other words, democracies don't fight each other, but does that have anything to do with democracy?


### What Are the Consequences of War?

War sucks and it creates costs, but there's still some major disagreement:

- Effects can be permanent and temporary, short-term and long-term, direct and indirect, positive and negative.

Discussions of "Phoenix" factors compound what we can say about war's lasting consequences.

# Conclusion
### Conclusion

We've done a lot of work on the causes of war.

- We have lots of answers.
- We generally don't have a single answer.

I enjoyed this class; I hope you did too.

- We'll discuss your final exam next.