---
title: "Power and the Realisms"
subtitle: POSC 3610  -- International Conflict
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: true
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```


```{r loadstuff, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
options(knitr.kable.NA = '')
library(car)
library(tidyverse)
library(janitor)
library(knitr)
library(stevemisc)
```

```{r loaddata, cache=T, eval=T, echo=F, message=F, error=F, warning=F}

# ICOWactive <- read.csv("~/Dropbox/data/icow/icow-provisional-1.01/200199.csv") %>% tbl_df()
CINC <- read.csv("~/Dropbox/data/cow/cinc/NMC_5_0.csv") %>% tbl_df()


```

# Introduction
### Goal for Today

*Discuss power as structural property and the various realism paradigms surrounding it.*

### What is Power?

Two conceptualizations of power focus on:

- relations
- resources

### Power as Relational

Common argument is that power is some kind of coercion.

- i.e. the ability to get someone else to do what they would otherwise not do.

Various aspects to power in this framework.

- Persuasion
- Rewards
- Punishments
- Coercion
- Generally: force-price-legitimacy framework

### Problems With This Interpretation

Several problems follow this concept of power for our purposes.

- Counterfactuals are hard
- Unobservables
- Attribution

### Power As Resources

More common interpretation in IR: power is resources.

- Major advantage: not conflating "power" (i.e. the cause) with outcomes we want to study (i.e. the effect)

# Power, Realism, and its Flaws
## Measuring Power
### Elements of Power

Any number of ways of measuring power (e.g. (in)tangible, observable/latent). Practically we go for:

- Terrain
- Natural resources (e.g. oil)
- Industrial capacity
- Military quality/preparedness
- Population
- Wealth (latent)
- National character (largely unobservable/stereotypes)

### Measuring Power

CoW's National Military Capabilities (NMC) data offer a crude measure of this concept of power.

$$
CINC_{it} = \frac{tpr_{it} + upr_{it} + ispr_{it} + ecr_{it} + mer_{it} + mpr_{it}}{6}
$$

..where:

- $tpr_{it}$ = total population ratio of country *i* in year *t*
- $upr_{it}$ = total urban population ratio of country *i* in year *t*
- $ispr_{it}$ = iron and steel production ratio of country *i* in year *t*
- $ecr_{it}$ = primary energy consumption ratio of country *i* in year *t*
- $mer_{it}$ = military expenditure ratio of country *i* in year *t*
- $mpr_{it}$ = military personnel ratio of country *i* in year *t*

###

```{r cinc-scores-usa-ukg-gmy-rus-1816-2010, echo=F, eval=T, fig.width = 14, fig.height = 8.5}

CINC %>%
  mutate(ccode = ifelse(ccode == 260, 255, ccode)) %>%
  filter(ccode == 2 | ccode == 200 | ccode == 255 | ccode == 365) %>%
  ggplot(., aes(year, cinc, group=factor(ccode), color=factor(ccode))) + geom_line(size=1.5) +
  theme_steve_web() +
  scale_x_continuous(breaks=seq(1820, 2010, by=10)) +
  xlab("Year") + ylab("CINC Score") + 
  labs(color = "Country",
       subtitle="The U.S. has long been the most powerful country in the world, but notice the various power transitions.",
       title="CINC Scores for the U.S., UK, Germany and Russia, 1816-2010",
       caption="Source: Correlates of War National Military Capabilities Data (v. 5.0)") +
  scale_colour_discrete(name="Country", 
                          labels=c("United States", "United Kingdom", "Germany/GFR",
                                   "Russia/USSR")) +
  theme(legend.position = "bottom")


```

## The Realisms
### Power as Structural Cause

We focus on the distribution of power in the international system because long-running paradigms are built around it.


### Classical Realism

Drawn from Hans Morgenthau's *Politics Among Nations*.

- Heavily inspired by Thomas Hobbes' *Leviathan*.
	- Anarchy reduces "Man" to his "nature".
- The state, *viz*, "Man" is hardwired to will for power.
- End result: bellum omnium contra omnes (war of all against all)

The state (i.e. "Man") pursues power to dominate his rivals.

- Nothing can be done to avoid this.

### Neorealism

Neorealism (aka "structural realism") remains the most prominent approach in security studies. The argument:

- The *structure* of the international system, not "human nature", forces states to pursue power.
- Anarchy has a single logic that forces a state to see means to protect itself.
- Power is the *means*, not the end.

### Neorealism's Assumptions

Neorealism is built on a few core assumptions (think: parsimony).

1. The international system is anarchic.
2. All states possess some type of offensive military capability.
3. States can never be 100% certain of other states' offensive intentions.
4. States are motivated to *survive*.
5. States are rational/strategic actors.

These assumptions will differ slightly from argument to argument.

- They actually come from Mearsheimer (2001).
- Most neorealist scholarship has done a poor job outlining its assumptions, as we shall see.

### Neorealism's Main Conclusions

All told, these assumptions imply states seek a **balance of power** in the international system.

- States eventually fear each other.
- This fear can never be inconsequential.
- International politics becomes a self-help world under anarchy.
- Power becomes the means to security.

Power-seeking leads to the famous problem of the **security dilemma**.

### Neorealism's Hypotheses

Several hypotheses follow these arguments.

- Bipolar systems are more stable than multipolar systems.
- States engage in balancing behavior, such that power distributions converge on a balance.
- States mimic, or echo, one another's behavior.

As we will see, these explanations are flawed in multiple ways.

- The assumptions do not logically imply the hypotheses.
- The empirical record does not vindicate the hypotheses.

## Bipolarity and Stability
### Bipolarity and Stability

Polarity constitutes possibly *the* core argument of neorealism:

- Bipolarity: peace
    - Reasons: certainty
- Multipolarity: war
    - Reasons: uncertainty. 
    - More specifically: **buck-passing** and **chain-ganging**

International system was multipolar before the Cold War

- The period saw multiple systemic wars dating back to 1648.
- Cold War was only point in history in which the two largest powers did not (directly) fight each other.

### Problems with the Polarity-Stability Hypothesis

- Not implied by any of the assumptions
- There was nothing special about the "long peace."

### The Hypotheses Do Not Follow the Assumptions

By itself, neorealism's assumptions do not imply the relationship between polarity and stability.

- i.e. "certainty" may embolden risk-taking, "uncertainty" may foster risk-aversion.
- We'd have to add another assumption: all states are equally risk-averse in the face of certainty.

If we relax this even a little bit, we've violated core assumptions of neorealism.

- Violates the unitary actor assumption
- Reduces hypothesized effect of polarity on stability to zero.
- States no longer mimic each other.

### The Polarity-Stability Relationship

Consider a world with A and B in which there are 300 units of "power".

- A: 150
- B: 150

Such a bipolar system would be stable.

- Neither A nor B could destroy each other.

### The Polarity-Stability Relationship

Consider a different world with A and B with 300 units of power.

- A: 151
- B: 149

Neorealism assumes this should be stable, but A could destroy B.

- Only when power is perfectly balanced does bipolarity produce peace.

Objection: power is balanced "enough".

- However, this would deny neorealism's own claim. Bipolarity is supposed to reduce uncertainty!

### The Polarity-Stability Relationship

Consider a five-country system as follows (with 300 units of power).

- A: 75
- B: 74
- C: 75
- D: 74
- E: 2

This system is incidentally stable.

- No one can be eliminated, not even E.

### Bipolarity, Uncertainty, and Stability

Can we salvage the bipolarity-stability argument if we relax the "uncertainty" claim?

- After all, our simple example may not do justice to understanding the real world.

Assume A thinks there's chance *p* it could eliminate B.

- *p* = A's resources/(B's resources + A's resources)

A does not attack B if:

\[ p(U_{AW}) + (1 - p)(U_{AL}) < U_{ASQ} \]

...where $U_{AW}$ = utility for A winning and $U_{AL}$ = utility for A losing.

### Bipolarity, Uncertainty, and Stability

Assume $U_{AW}$ = 1 and $U_{AL}$ = 0. When would A attack B?

\begin{eqnarray}
p(U_{AW}) + (1 - p)(U_{AL}) &>& U_{ASQ}   \nonumber \\
pU_{AW} + U_{AL} - pU_{AL}  &>& U_{ASQ} \nonumber \\
pU_{AW}  - pU_{AL}  &>& U_{ASQ} - U_{AL} \nonumber \\
p   &>& \frac{U_{ASQ} - U_{AL}}{U_{AW}  - U_{AL}} \nonumber \\
p   &>& \frac{U_{ASQ} - 0}{1 - 0} \nonumber \\
p  &>& U_{ASQ} \nonumber
\end{eqnarray}

A attacks B if the probability of winning is greater than A's utility of the status quo.

### Bipolarity, Uncertainty, and Stability

Assume a world of 300 units of power.

- A: 60
- B: 240

When would A attack B?

- $p = \frac{60}{60+240} = .2$
- If A is really dissatisfied with the status quo (i.e. $U_{ASQ} < .2$), it'll attack B.

This is intuitive but it violates a neorealist assumption of security-oriented behavior!

###

```{r stability-international-systems-1492-1990, echo=F, eval=T, fig.width = 14, fig.height = 8.5}
tibble(
  year = c(1492, 1519, 1556, 1609, 1617, 1699, 1713, 1721, 1740, 1808,
           1861, 1898, 1905, 1918, 1943, 1945, 1949),
  length = c(1519-1492,
             1556-1519,
             1609-1556,
             1617-1609,
             1699-1617,
             1713-1699,
             1721-1713,
             1740-1721,
             1808-1740,
             1861-1808,
             1898-1861,
             1905-1898,
             1918-1905,
             1943-1918,
             1945-1943,
             1949-1945,
             1990-1949)
) %>%
  ggplot(.,aes(as.factor(year), length)) + 
  geom_bar(stat="identity", alpha=0.8, color="black", fill="#619cff") + 
  theme_steve_web() +
  geom_text(aes(label=length), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  xlab("Year of Change in List of Major Powers") + ylab("Duration (in Years) of Era") +
  ggtitle("Stability of International Systems (1492-1990)") +
  labs(subtitle="i.e. there is nothing special about the ''Long Peace'' of the Cold War.",
       caption = "Source: Bueno de Mesquita (2010). Note: 'Stability' defined as a change in the composition of major powers.")

```

# Conclusion
### Conclusion

We study power because we believes its distribution matters to war and peace.

- For our purposes, better to focus on resources than relational power.


Neorealism purports to be a parsimonious explanation of international politics.

- It's also the most common approach in security studies.

However, neorealism suffers from major flaws.

- The assumptions do not imply the hypotheses.
- The hypotheses, however derived, are not supported by the empirical record.