---
title: "The Democratic Peace"
subtitle: POSC 3610  -- International Conflict
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: false
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
options(knitr.kable.NA = '')
```

```{r loadstuff, include=FALSE, eval=T}
knitr::opts_chunk$set(cache=FALSE)


library(tidyverse)
library(stevemisc)
library(knitr)
library(dotwhisker)
library(artyfarty)
# library(QuickUDS)

Polity <- readxl::read_xls("~/Dropbox/data/polity/p4v2016.xls")
Dat <- read.csv("~/Dropbox/projects/mid-project/gml-mid-data/2.1/gml-ndy-2.1.csv") %>% tbl_df()

Polity %>%
  select(ccode, scode, year, polity2) %>%
      mutate(ccode = ifelse(ccode == 305 & year < 1919, 300, ccode),
             ccode = ifelse(ccode == 347, 345, ccode),
             ccode = ifelse(ccode == 364, 365, ccode),
             ccode = ifelse(ccode == 769, 770, ccode),
             ccode = ifelse(ccode == 818, 816, ccode),
             ccode = ifelse(ccode == 529, 530, ccode)) %>%
  rename(ccode1 = ccode,
         scode1 = scode,
         polity21 = polity2) %>%
  left_join(Dat, .) -> Dat

Polity %>%
  select(ccode, scode, year, polity2) %>%
      mutate(ccode = ifelse(ccode == 305 & year < 1919, 300, ccode),
             ccode = ifelse(ccode == 347, 345, ccode),
             ccode = ifelse(ccode == 364, 365, ccode),
             ccode = ifelse(ccode == 769, 770, ccode),
             ccode = ifelse(ccode == 818, 816, ccode),
             ccode = ifelse(ccode == 529, 530, ccode)) %>%
  rename(ccode2 = ccode,
         scode2 = scode,
         polity22 = polity2) %>%
  left_join(Dat, .) -> Dat

Dat %>% 
  mutate(war = ifelse(hostlev == 5, 1, 0),
         bilat = ifelse(numa + numb == 2, 1, 0),
         jointdem = ifelse(polity21 >= 6 & polity22 >= 6, 1, 0),
         onedem = ifelse((polity21 >=6 & polity22 < 6) | (polity21 < 6 & polity22 >= 6), 1, 0),
         nodem = ifelse(polity21 < 6 & polity22 < 6, 1, 0),
         demcat = NA,
         demcat = ifelse(jointdem == 1, "Joint Democracy", demcat),
         demcat = ifelse(onedem == 1, "One Side Democratic", demcat),
         demcat = ifelse(nodem == 1, "No Side Democratic", demcat)) -> Dat


```

```{r data, include=FALSE, eval=F}

Data <- read.csv("~/Dropbox/projects/mid-project/gml-mid-data/2.1/gml-ndy-2.1.csv") %>% tbl_df()
Part <- read.csv("~/Dropbox/projects/mid-project/gml-mid-data/2.1/gml-midb-2.1.csv") %>% tbl_df()
States <- read.csv("~/Dropbox/data/cow/states/states2016.csv") %>% tbl_df()

Cont <- read.csv("~/Dropbox/data/cow/contiguity/3.2/contdird.csv") %>% tbl_df()
Majors <- read.csv("~/Dropbox/data/cow/states/majors2016.csv") %>% tbl_df()
Contcd <- read.csv("~/Dropbox/data/cow/coldepcont/contcold.csv") %>% tbl_df()
Alliance <- read_csv("~/Dropbox/data/cow/alliance/4.1/alliance_v4.1_by_dyad_yearly.csv")
NMC <- read_csv("~/Dropbox/data/cow/cinc/NMC_5_0.csv")
Polity <- readxl::read_xls("~/Dropbox/data/polity/p4v2016.xls")
ATOP <- haven::read_dta("~/Dropbox/data/atop/atop3_0ddyr.dta")
IGO <- read.csv("~/Dropbox/data/cow/igo/IGO_dyadunit_v2.3.csv", na.string=c("-1", "-9")) %>% tbl_df()
Trade <- read.csv("~/Dropbox/data/cow/trade/Dyadic_COW_4.0.csv", na.string=c("-9")) %>% tbl_df()

Contcd %>%
  rename(ccode1 = statelno, ccode2 = statehno) %>%
  select(ccode1, ccode2, year, land, sea, total) %>%
  left_join(Data, .) -> Data

Cont %>%
  rename(ccode1 = state1no, ccode2 = state2no) %>%
  select(ccode1, ccode2, year, conttype) %>%
  left_join(Data, .) -> Data

Data %>%
  mutate(conttype = ifelse(is.na(conttype), 6, conttype),
         land = ifelse(is.na(land), 0, land),
         sea = ifelse(is.na(sea), 0, sea),
         total = ifelse(is.na( total), 0, total),
         contig = ifelse(conttype == 6 & total == 0, 0, 1),
         # Code regions...
         region1 = NA,
         region1 = ifelse(ccode1 < 200, "Americas", region1),
         region1 = ifelse(ccode1 >= 200 & ccode1 < 400, "Europe", region1),
         region1 = ifelse(ccode1 >= 400 & ccode1 < 600, "Africa", region1),
         region1 = ifelse(ccode1 >= 600 & ccode1 < 700, "ME/NA", region1),
         region1 = ifelse(ccode1 >= 700 & ccode1 < 900, "Asia", region1),
         region1 = ifelse(ccode1 >= 900, "Oceania", region1),
         region2 = NA,
         region2 = ifelse(ccode2 < 200, "Americas", region2),
         region2 = ifelse(ccode2 >= 200 & ccode2 < 400, "Europe", region2),
         region2 = ifelse(ccode2 >= 400 & ccode2 < 600, "Africa", region2),
         region2 = ifelse(ccode2 >= 600 & ccode2 < 700, "ME/NA", region2),
         region2 = ifelse(ccode2 >= 700 & ccode2 < 900, "Asia", region2),
         region2 = ifelse(ccode2 >= 900, "Oceania", region2),
         # Code majors...
         maj1 = NA,
         maj1 = ifelse(ccode1 == 2 & year >= 1898, 1, maj1),
         maj1 = ifelse(ccode1 == 200, 1, maj1),
         maj1 = ifelse(ccode1 == 220 & year <= 1940, 1, maj1),
         maj1 = ifelse(ccode1 == 220 & year >= 1945, 1, maj1),
         maj1 = ifelse(ccode1 == 255 & year <= 1918, 1, maj1),
         maj1 = ifelse(ccode1 == 255 & (year >= 1925 & year <= 1945), 1, maj1),
         maj1 = ifelse(ccode1 == 255 & year >= 1991, 1, maj1),
         maj1 = ifelse(ccode1 == 300 & year <= 1918, 1, maj1),
         maj1 = ifelse(ccode1 == 325 & (year >= 1860 & year <= 1943), 1, maj1),
         maj1 = ifelse(ccode1 == 365 & year <= 1917, 1, maj1),
         maj1 = ifelse(ccode1 == 365 & year >= 1922, 1, maj1),
         maj1 = ifelse(ccode1 == 710 & year >= 1950, 1, maj1),
         maj1 = ifelse(ccode1 == 740 & (year >= 1895 & year <= 1945), 1, maj1),
         maj1 = ifelse(ccode1 == 740 & year >= 1991, 1, maj1),
         maj1 = ifelse(is.na(maj1), 0, maj1),
         maj2 = NA,
         maj2 = ifelse(ccode2 == 2 & year >= 1898, 1, maj2),
         maj2 = ifelse(ccode2 == 200, 1, maj2),
         maj2 = ifelse(ccode2 == 220 & year <= 1940, 1, maj2),
         maj2 = ifelse(ccode2 == 220 & year >= 1945, 1, maj2),
         maj2 = ifelse(ccode2 == 255 & year <= 1918, 1, maj2),
         maj2 = ifelse(ccode2 == 255 & (year >= 1925 & year <= 1945), 1, maj2),
         maj2 = ifelse(ccode2 == 255 & year >= 1991, 1, maj2),
         maj2 = ifelse(ccode2 == 300 & year <= 1918, 1, maj2),
         maj2 = ifelse(ccode2 == 325 & (year >= 1860 & year <= 1943), 1, maj2),
         maj2 = ifelse(ccode2 == 365 & year <= 1917, 1, maj2),
         maj2 = ifelse(ccode2 == 365 & year >= 1922, 1, maj2),
         maj2 = ifelse(ccode2 == 710 & year >= 1950, 1, maj2),
         maj2 = ifelse(ccode2 == 740 & (year >= 1895 & year <= 1945), 1, maj2),
         maj2 = ifelse(ccode2 == 740 & year >= 1991, 1, maj2),
         maj2 = ifelse(is.na(maj2), 0, maj2)) -> Data

Data %>%
  mutate(region = NA,
         region = ifelse(region1 == "Americas" & region2 == "Americas", "Americas", region),
         region = ifelse(region1 == "Europe" & region2 == "Europe", "Europe", region),
         region = ifelse(region1 == "Africa" & region2 == "Africa", "Africa", region),
         region = ifelse(region1 == "ME/NA" & region2 == "ME/NA", "ME/NA", region),
         region = ifelse(region1 == "Asia" & region2 == "Asia", "Asia", region),
         region = ifelse(region1 == "Oceania" & region2 == "Oceania", "Oceania", region),
         prd = NA,
         prd = ifelse(contig == 1 | (maj1 == 1 | maj2 == 1), 1, 0)) -> Data


Alliance %>%
  select(ccode1, ccode2, year, defense, neutrality, nonaggression, entente) %>%
  left_join(Data, .) %>%
  mutate(defense = ifelse(is.na(defense), 0, defense),
         neutrality = ifelse(is.na(neutrality), 0, neutrality),
         nonaggression = ifelse(is.na(nonaggression), 0, nonaggression),
         entente = ifelse(is.na(entente), 0, entente)) -> Data

NMC %>%
  select(ccode:cinc) %>%
  rename(ccode1 = ccode,
         milex1 = milex,
         milper1 = milper,
         irst1 = irst,
         pec1 = pec,
         tpop1 = tpop,
         upop1 = upop,
         cinc1 = cinc) %>%
  left_join(Data, .) -> Data

NMC %>%
  select(ccode:cinc) %>%
  rename(ccode2 = ccode,
         milex2 = milex,
         milper2 = milper,
         irst2 = irst,
         pec2 = pec,
         tpop2 = tpop,
         upop2 = upop,
         cinc2 = cinc) %>%
  left_join(Data, .) -> Data

#democracy %>% tbl_df() %>%
#  select(cown, year, polity2) %>%
#  rename(ccode1 = cown,
#         polity21 = polity2) %>%
#  left_join(NDY, .)

Polity %>%
  select(ccode, scode, year, polity2) %>%
      mutate(ccode = ifelse(ccode == 305 & year < 1919, 300, ccode),
             ccode = ifelse(ccode == 347, 345, ccode),
             ccode = ifelse(ccode == 364, 365, ccode),
             ccode = ifelse(ccode == 769, 770, ccode),
             ccode = ifelse(ccode == 818, 816, ccode),
             ccode = ifelse(ccode == 529, 530, ccode)) %>%
  rename(ccode1 = ccode,
         scode1 = scode,
         polity21 = polity2) %>%
  left_join(Data, .) -> Data

Polity %>%
  select(ccode, scode, year, polity2) %>%
      mutate(ccode = ifelse(ccode == 305 & year < 1919, 300, ccode),
             ccode = ifelse(ccode == 347, 345, ccode),
             ccode = ifelse(ccode == 364, 365, ccode),
             ccode = ifelse(ccode == 769, 770, ccode),
             ccode = ifelse(ccode == 818, 816, ccode),
             ccode = ifelse(ccode == 529, 530, ccode)) %>%
  rename(ccode2 = ccode,
         scode2 = scode,
         polity22 = polity2) %>%
  left_join(Data, .) -> Data

Data %>% 
  mutate(war = ifelse(hostlev == 5, 1, 0),
         landcontig = ifelse(conttype == 1, 1, 0),
         othercontig = ifelse(conttype > 1 & conttype != 0, 1, 0),
         bilat = ifelse(numa + numb == 2, 1, 0),
         allied = ifelse(defense == 1 | neutrality == 1 | nonaggression == 1 | entente == 1, 1, 0),
         jointdem = ifelse(polity21 >= 6 & polity22 >= 6, 1, 0),
         relpow = ifelse(cinc1 > cinc2, cinc1/cinc2, cinc2/cinc1),
         relpowd = ifelse(relpow >=3, 1, 0)) -> Data

Data %>%
  mutate(dyad = as.numeric(paste0("1",sprintf("%03d", ccode1), 
                                    sprintf("%03d", ccode2))),
         fatal = ifelse(fatality > 0, 1, 0),
         fatmid = ifelse(fatal == 1 & midongoing == 1, 1, 0),
         fatmidonset = ifelse(fatmid == 1 & midonset == 1, 1, 0)) -> Data

Data <- sbtscs(Data, midongoing, year, dyad) %>% tbl_df() %>%
  rename(midspell = spell)
Data <- sbtscs(Data, fatmid, year, dyad) %>% tbl_df() %>%
  rename(fatspell = spell)

IGO %>%
  select(ccode1, ccode2, year, AAAID:Wassen) %>%
  replace(is.na(.), 0) %>% 
  mutate(numigo = rowSums(.[4:ncol(.)])) %>%
  select(ccode1, ccode2, year, numigo) %>%
  left_join(Data, .) %>%
  fill(numigo) -> Data

Trade %>%
  select(ccode1, ccode2, year, smoothtotrade) %>%
  left_join(Data, .) -> Data

```

# Introduction
### Goal for Today

*Discuss the democratic peace and its importance to the study of conflict.*

###

```{r polity, echo=F, eval=T, fig.width = 14, fig.height = 8.5, warning = F, message = F}

Polity %>% 
  mutate(dem = ifelse(polity2 >= 6, 1, 0)) %>%
  group_by(year) %>%
  summarize(demrate = mean(dem, na.rm=T)) %>%
  ggplot(., aes(year, demrate)) + geom_line(size=1) + theme_steve_web() +
  scale_y_continuous(labels=scales::percent, limits=c(0,1)) +
  xlab("Year") + ylab("Percentage of International System That's Democratic") +
  scale_x_continuous(breaks = seq(1800, 2020, by = 10)) +
  geom_ribbon(aes(ymin=0, ymax=demrate),
              alpha=0.3, fill="blue") +
  labs(caption="Data: The Polity Project (Center for Systemic Peace)",
       subtitle="States coded with `polity2` value at or above 6 are coded as democratic.",
       title="The Proliferation of Democracies in the International System (1800-2016)")

```

###

```{r demrates, echo=F, eval=T, fig.width = 14, fig.height = 8.5, warning = F, message = F}
Cont <- read.csv("~/Dropbox/data/cow/contiguity/3.2/contdird.csv") %>% tbl_df() %>%
  rename(ccode1 = state1no, ccode2 = state2no) %>%
  select(ccode1, ccode2, year, conttype)

Polity %>% 
  mutate(dem = ifelse(polity2 >= 6, 1, 0)) %>%
  group_by(year) %>%
  summarize(`Proportion of Democracies` = mean(dem, na.rm=T)) -> dems


Dat %>% 
  left_join(., Cont) %>%
  filter(!is.na(conttype)) %>%
  group_by(year) %>%
  summarize(`Proportion of States in a MID` = mean(midongoing, na.rm=T)) %>%
  left_join(dems, .) %>%
  gather(Category, Value, `Proportion of Democracies`:`Proportion of States in a MID`) %>%
  filter(year >= 1816 & year <= 2010) %>%
  ggplot(.,aes(year, Value, group=Category, color=Category, linetype=Category)) + theme_steve_web() +
  geom_line(size=1.1) +
  scale_color_manual(values = pal("google")) +
  scale_x_continuous(breaks = seq(1820, 2010, by = 10)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Year") + ylab("Percentage") +
  labs(title = "The International System Has Seen Less Conflict as It Became More Democratic",
       subtitle = "Admittedly, most of the relationship is observable after WWII and may cluster on something else.",
       caption = "Data: GML MID Data (v. 2.1) and the Polity Project")
```

# Democratic Peace Theory
### Democratic Peace Theory

Democratic peace theory (DPT) may constitute the most important advancement in IR scholarship.

- Originally a finding by Babst (1964), a skeptical Singer and Small (1976) confirmed it.
- Has important theoretical origins in Immanuel Kant (1795).
- Levy (1988) notes it's the closest thing to an empirical law in all political science.

Long story short: democracies don't fight each other, and never in war.

### Democratic Peace Theory

Nonetheless, DPT is a bit of a misnomer.

- It's a democratic peace *fact.* It *needs* a theory.
- In many ways, this is still true.

So, why don't democracies fight each other?

## The Arguments
### Maoz and Russett (1993)

Maoz and Russett (1993) test two competing explanations.


- Normative model
- Structural (institutional) model

### Normative Model

The authors identify two assumptions of the normative model. 

1. States externalize their *internal* norms of behavior.
2. A conflict between democrats and autocrats will be characterized by the norms of the latter.

### Normative Model

Autocrats have few if any normative bounds on their behavior.

- They may capture the state through use of lethal force and violence.
- They may also keep their hold on power through the same means.
- Autocratic foreign policy behavior reflects the autocrat's preferences.

### Normative Model

Democracies, by contrast, rest on different norms.

- e.g. equal competition, minority rights, consent to be governed.
- Force and repression to govern would be deemed "illegitimate."
- This imposes "normative" restraints on behavior.

International politics becomes an extension of domestic politics.

### The Normative Model

The argument:

- In a jointly democratic dispute, both sides are secure in their knowledge of the other's normative restraints.
- In a mixed or autocratic dispute, nothing is in place to restrain escalation.

Democracies will not fight each other, but will fight other pairs of states.


### Structural (Institutional) Model

The authors identify two assumptions of the structural model.

1. Dangerous foreign policy dilemmas require a lengthy mobilization of domestic support.
2. Only emergencies allow democracies to circumvent this mobilization process.


### Structural (Institutional) Model

The argument:

- In a jointly democratic dispute, selling the conflict domestically takes too much time.
    - Cooler heads will prevail.
- There are no mutual structural constraints in a mixed or autocratic dispute.
    - This resembles an "emergency" that will allow democratic leaders to circumvent the lengthy mobilization process.

Democracies will not fight each other, but will fight other pairs of states.

## 
### The Results

```{r, echo=F, eval=T}
Dat %>%
  group_by(jointdem) %>%
  filter(!is.na(jointdem)) %>%
  filter(midonset == 1) %>%
  summarize(n = sum(war)) %>%
  mutate(jointdem = ifelse(jointdem == 0, "No Joint Democracy", jointdem),
         jointdem = ifelse(jointdem == 1, "Joint Democracy", jointdem)) %>%
  rename(`**Joint Democracy?**` = jointdem,
         `**Number of Wars**` = n) %>%
  kable(., caption = "Simple Bivariate Relationship Between Joint Democracy and War (GML MID Non-Directed Dyad-Year Data [v. 2.1]",
        align = c("l", "c"))

```

### So What Are Those Joint Democratic Wars?

Flukes, basically:

- Coding artifact of ephemeral declaration of war from WWII allies to Finland (MID#0258)
- Franco-Siamese War (MID#0196) even though Britain never actually fought France.
- India-Pakistan "Kargil War" (MID#4007) even though Pakistan was reverting to a police state
- Cyprus-Turkey 1974 conflict (MID#1293) even though Turkey was relapsing to authoritarianism again.

### A More Systematic Analysis

Here's a more rigorous dyad-year analysis:

- *DV*: MID ocurrence, MID initiation, MID escalation
- *IV*: joint democracy, only one side democratic.
    - We include both to test for a "cats-and-dogs" effect.
    - i.e. we expect joint democracy to have a negative effect and one-side-democratic to have a positive effect.
- *Controls:* distance, power parity, peace years, *S*-score for affinity in alliances.

###

```{r simple-dp-models, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}


tribble(
  ~term, ~estimate, ~std.error, ~model,
  "Both Democratic", -.3527, .1279, "Dispute Occurrence",
  "One Side Democratic", .4586, .0776, "Dispute Occurrence",
  "Alliance Affinity", -.2747, .2330, "Dispute Occurrence",
  "ln(Distance)", -.2469, .0131, "Dispute Occurrence",
  "Power Parity", .0477, .1514, "Dispute Occurrence",
  "Peace Years", -.2979, .0171, "Dispute Occurrence"
) %>%
  mutate(statistic = estimate/std.error,
         p.value = 1.96*pnorm(-abs(statistic))) %>%
  dwplot(.,dot_args = list(aes(colour = model, shape = model), size = 1.1)) + theme_steve_web() +
  xlab("Coefficient Estimate") +
  geom_vline(xintercept = 0, colour = "grey60", linetype = 2) +
  theme(legend.position = "bottom") +
  labs(title = "Democracies are Peaceful, but Just With Each Other",
       subtitle = "Jointly democratic dyads are less likely to be in a MID but dyads in which just one side is democratic is more conflict-prone than jointly autocratic dyads.",
       caption = "Data: Quackenbush (2015).") +
  scale_colour_discrete(name = "Model") + scale_shape_discrete(name = "Model")

```

###

```{r dp-models-with-interactions, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}

tribble(
  ~term, ~estimate, ~std.error, ~model,
  "State A is Democratic", -.1213, .0842, "Model 1 (No Interaction)",
  "State B is Democratic", .1821, .0760, "Model 1 (No Interaction)",
  "Joint Democracy", NA, NA, "Model 1 (No Interaction)",
  "Alliance Affinity", -.4162, .2363, "Model 1 (No Interaction)",
  "ln(Distance)", -.2627, .0139, "Model 1 (No Interaction)",
  "Relative Power", .6359, .0865, "Model 1 (No Interaction)",
  "Peace Years", -.2787, .0180, "Model 1 (No Interaction)",
  "State A is Democratic", .3499, .0985, "Model 2 (Joint Democracy Interaction)",
  "State B is Democratic", .5954, .0905, "Model 2 (Joint Democracy Interaction)",
  "Joint Democracy", -1.2622, .1522, "Model 2 (Joint Democracy Interaction)",
  "Alliance Affinity", -.3162, .2365, "Model 2 (Joint Democracy Interaction)",
  "ln(Distance)", -.2729, .0162, "Model 2 (Joint Democracy Interaction)",
  "Relative Power", .6326, .0860, "Model 2 (Joint Democracy Interaction)",
  "Peace Years", -.2758, .0178, "Model 2 (Joint Democracy Interaction)"
) %>%
  mutate(statistic = estimate/std.error,
         p.value = 1.96*pnorm(-abs(statistic))) %>%
  dwplot(.,dot_args = list(aes(colour = model, shape = model), size = 1.1)) + theme_steve_web() +
  xlab("Coefficient Estimate") +
  geom_vline(xintercept = 0, colour = "grey60", linetype = 2) +
  theme(legend.position = "bottom") +
  labs(title = "Democracies Appear to be Conflict Prone, but Just Not With Each Other",
       subtitle = "Jointly democratic dyads are less likely to be in a MID but dyads in which just one side is democratic is more conflict-prone than jointly autocratic dyads.",
       caption = "Data: Quackenbush (2015).") +
  scale_colour_discrete(name = "Model") + scale_shape_discrete(name = "Model")

```


###

```{r monadic-dp, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}
tribble(
  ~term, ~estimate, ~std.error, ~model,
  "Democracy", .0109, .0842, "MID Involvement",
  "ln(Power)", .2860, .0326, "MID Involvement",
  "Democracy", -.0042, .1647, "MID Initiation",
  "ln(Power)", .3479, .0347, "MID Initiation"
) %>%
  mutate(statistic = estimate/std.error,
         p.value = 1.96*pnorm(-abs(statistic))) %>%
  dwplot(.,dot_args = list(aes(colour = model, shape = model), size = 1.1)) + theme_steve_web() +
  xlab("Coefficient Estimate") +
  geom_vline(xintercept = 0, colour = "grey60", linetype = 2) +
  theme(legend.position = "bottom") +
  labs(title = "There Is No Monadic Democratic Peace",
       subtitle = "Democracies are no different in their proclivity for conflict and almost all models have failed to vindicate the monadic proposition.",
       caption = "Data: Quackenbush (2015).") +
  scale_colour_discrete(name = "Model") + scale_shape_discrete(name = "Model")
```

### Other Findings and Arguments

Democratic peace findings generally cluster into two categories:

1. Democratic war avoidance
2. Democratic war behavior

### Democratic War Avoidance

1. Democracies don't fight wars against each other.
2. Democra*tizing* states are war-prone (e.g. Serbia). Mature democracies are the peaceful ones.
3. Democracies conclude what disputes they do have with negotiation and compromise.
4. Democracies as major powers are more constrained than democratic minor powers.

### Democratic War Behavior

1. Democracies are essentially as war-prone as non-democracies.
2. Democracies win the wars they fight.
3. Democracies fight shorter wars.
4. Democracies are as likely as non-democracies to target weak rivals.
5. Democracies are more likely to initiate wars against autocracies (not the other way around).
6. Democracies incur fewer battle deaths in the wars they initiate.

### Auxiliary Arguments

Others have added some tweaks to the core arguments.

- "Kantian peace"
- "Selectorate peace"

Neither seriously challenges that democracy is a force for peace, just that there's more happening than the standard DPT argument.

### 

```{r, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}
Data <- read.csv("~/Dropbox/projects/mid-project/gml-mid-data/2.02/gml-ndy-2.02.csv") %>% tbl_df()
Cont <- read.csv("~/Dropbox/data/cow/contiguity/3.2/contdird.csv") %>% tbl_df()
Majors <- read.csv("~/Dropbox/data/cow/states/majors2016.csv") %>% tbl_df()
Contcd <- read.csv("~/Dropbox/data/cow/coldepcont/contcold.csv") %>% tbl_df()
Trade <- read.csv("~/Dropbox/data/cow/trade/Dyadic_COW_4.0.csv") %>% tbl_df()

Trade %>% 
  select(ccode1, ccode2, year, smoothtotrade) %>%
  left_join(Data, .) -> Data

Contcd %>%
  rename(ccode1 = statelno, ccode2 = statehno) %>%
  select(ccode1, ccode2, year, land, sea, total) %>%
  left_join(Data, .) -> Data

Cont %>%
  rename(ccode1 = state1no, ccode2 = state2no) %>%
  select(ccode1, ccode2, year, conttype) %>%
  left_join(Data, .) -> Data

Data %>%
  mutate(conttype = ifelse(is.na(conttype), 6, conttype),
         land = ifelse(is.na(land), 0, land),
         sea = ifelse(is.na(sea), 0, sea),
         total = ifelse(is.na( total), 0, total),
         contig = ifelse(conttype == 6 & total == 0, 0, 1),
         # Code regions...
         region1 = NA,
         region1 = ifelse(ccode1 < 200, "Americas", region1),
         region1 = ifelse(ccode1 >= 200 & ccode1 < 400, "Europe", region1),
         region1 = ifelse(ccode1 >= 400 & ccode1 < 600, "Africa", region1),
         region1 = ifelse(ccode1 >= 600 & ccode1 < 700, "ME/NA", region1),
         region1 = ifelse(ccode1 >= 700 & ccode1 < 900, "Asia", region1),
         region1 = ifelse(ccode1 >= 900, "Oceania", region1),
         region2 = NA,
         region2 = ifelse(ccode2 < 200, "Americas", region2),
         region2 = ifelse(ccode2 >= 200 & ccode2 < 400, "Europe", region2),
         region2 = ifelse(ccode2 >= 400 & ccode2 < 600, "Africa", region2),
         region2 = ifelse(ccode2 >= 600 & ccode2 < 700, "ME/NA", region2),
         region2 = ifelse(ccode2 >= 700 & ccode2 < 900, "Asia", region2),
         region2 = ifelse(ccode2 >= 900, "Oceania", region2),
         # Code majors...
         maj1 = NA,
         maj1 = ifelse(ccode1 == 2 & year >= 1898, 1, maj1),
         maj1 = ifelse(ccode1 == 200, 1, maj1),
         maj1 = ifelse(ccode1 == 220 & year <= 1940, 1, maj1),
         maj1 = ifelse(ccode1 == 220 & year >= 1945, 1, maj1),
         maj1 = ifelse(ccode1 == 255 & year <= 1918, 1, maj1),
         maj1 = ifelse(ccode1 == 255 & (year >= 1925 & year <= 1945), 1, maj1),
         maj1 = ifelse(ccode1 == 255 & year >= 1991, 1, maj1),
         maj1 = ifelse(ccode1 == 300 & year <= 1918, 1, maj1),
         maj1 = ifelse(ccode1 == 325 & (year >= 1860 & year <= 1943), 1, maj1),
         maj1 = ifelse(ccode1 == 365 & year <= 1917, 1, maj1),
         maj1 = ifelse(ccode1 == 365 & year >= 1922, 1, maj1),
         maj1 = ifelse(ccode1 == 710 & year >= 1950, 1, maj1),
         maj1 = ifelse(ccode1 == 740 & (year >= 1895 & year <= 1945), 1, maj1),
         maj1 = ifelse(ccode1 == 740 & year >= 1991, 1, maj1),
         maj1 = ifelse(is.na(maj1), 0, maj1),
         maj2 = NA,
         maj2 = ifelse(ccode2 == 2 & year >= 1898, 1, maj2),
         maj2 = ifelse(ccode2 == 200, 1, maj2),
         maj2 = ifelse(ccode2 == 220 & year <= 1940, 1, maj2),
         maj2 = ifelse(ccode2 == 220 & year >= 1945, 1, maj2),
         maj2 = ifelse(ccode2 == 255 & year <= 1918, 1, maj2),
         maj2 = ifelse(ccode2 == 255 & (year >= 1925 & year <= 1945), 1, maj2),
         maj2 = ifelse(ccode2 == 255 & year >= 1991, 1, maj2),
         maj2 = ifelse(ccode2 == 300 & year <= 1918, 1, maj2),
         maj2 = ifelse(ccode2 == 325 & (year >= 1860 & year <= 1943), 1, maj2),
         maj2 = ifelse(ccode2 == 365 & year <= 1917, 1, maj2),
         maj2 = ifelse(ccode2 == 365 & year >= 1922, 1, maj2),
         maj2 = ifelse(ccode2 == 710 & year >= 1950, 1, maj2),
         maj2 = ifelse(ccode2 == 740 & (year >= 1895 & year <= 1945), 1, maj2),
         maj2 = ifelse(ccode2 == 740 & year >= 1991, 1, maj2),
         maj2 = ifelse(is.na(maj2), 0, maj2)) -> Data

Data %>%
  mutate(region = NA,
         region = ifelse(region1 == "Americas" & region2 == "Americas", "Americas", region),
         region = ifelse(region1 == "Europe" & region2 == "Europe", "Europe", region),
         region = ifelse(region1 == "Africa" & region2 == "Africa", "Africa", region),
         region = ifelse(region1 == "ME/NA" & region2 == "ME/NA", "ME/NA", region),
         region = ifelse(region1 == "Asia" & region2 == "Asia", "Asia", region),
         region = ifelse(region1 == "Oceania" & region2 == "Oceania", "Oceania", region),
         prd = NA,
         prd = ifelse(contig == 1 | (maj1 == 1 | maj2 == 1), 1, 0)) -> Data

Data %>%
  filter(prd == 1) %>%
  mutate(dyad = as.numeric(paste0("1",sprintf("%03d", ccode1), 
                                    sprintf("%03d", ccode2))),
         logtrade = ifelse(smoothtotrade < 0, NA, log(smoothtotrade + 1))) -> PRD

PRD <- sbtscs(PRD, midongoing, year, dyad) %>% tbl_df()

require(brglm)
require(splines)
M1 <- brglm(midonset ~ logtrade + bs(spell, 4),
             data=PRD, family=binomial(logit), 
             na.action = na.exclude, method="brglm.fit", pl=TRUE)

newdat <- data.frame(logtrade = seq(min(PRD$logtrade, na.rm=TRUE), max(PRD$logtrade, na.rm=T), by = 0.01),
                     spell = 0)

predict(M1, newdat) -> preds

PRD <- cbind(PRD, pmid = predict(M1,type='response'))

PRD %>%
  group_by(year) %>%
  filter(year > 1869) %>%
  summarize(pmid = mean(midongoing, na.rm=T),
            meantrade = mean(logtrade, na.rm=T),
            rmeantrade = (meantrade - min(meantrade))/(max(meantrade) - min(meantrade))) %>%
  ggplot(.,aes(x = year)) +
  geom_line(aes(y = pmid, color = "Proportion of States in a MID"), size=1.1) +
  geom_line(aes(y = meantrade/6, color = "Total Volume of Trade (Logged)"), size=1.1) + 
  scale_y_continuous(sec.axis = sec_axis(~.*6, name = "Average Total Volume of Trade in the World (Logged)")) +
  theme_steve_web() + ylab("Proportion of States in a MID") + xlab("Year") +
  scale_x_continuous(breaks=seq(1870, 2010, by = 5)) +
  labs(colour = "Category",
       title = "IR Scholars Generally See Trade as Conducive to Global Peace",
       subtitle = "The argument is intuitive; interconnected states see conflict as too costly and an unnecessary disruption of commerce.",
       caption = "Data: Correlates of War Dyadic Trade Data (v. 4.0) and Gibler-Miller-Little MID Data (v. 2.02)")

```


# Conclusion
### Conclusion

Democratic peace might be the most important finding in IR, but doubts persist:

- "Capitalist peace" (ed. there are a laundry list of problems with this argument)
- Common systems or common interests? (ed. we don't adequately model temporal variation)
- Territorial peace (i.e. have we put the cart before the horse?)