---
output: 
  stevetemplates::beamer:
    latex_engine: xelatex # use xelatex here instead! I recommend it, but this is minimal reprex
    dev: cairo_pdf # I typically comment this out  if latex_engine: pdflatex
    slide_level: 3 # I prefer this, but I won't force it on you.
title: "Whither War? An Empirical Assessment of the Patterns of War"
subtitle: POSC 3610  -- International Conflict
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: "`r paste0(Sys.getenv('HOME'), '/Dropbox/clemson/watermarks/clemson-university-wordmark-orange-purple.png')`"
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=F,
                      fig.path = 'figs/',
                      cache.path='cache/',
                      warning=F,
                      message=F,
                      fig.width = 14, fig.height = 8.5,
                      echo=F)

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}

library(stevemisc)
library(tidyverse)
library(kableExtra)
library(peacesciencer)

getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}


mode <- function(codes){
  which.max(tabulate(codes))
}

Conf <- read_csv("~/Dropbox/projects/mid-project/internal/conf-diagnostic-check.csv") %>%
  mutate(is_war = ifelse(fatalmin >= 1000, "War", "Sub-War Confrontation"))
cow_extra <- read_csv("~/Dropbox/data/cow/wars/Extra-StateWarData_v4.0.csv") %>% rename_all(tolower)
cow_non <- read_csv("~/Dropbox/data/cow/wars/Non-StateWarData_v4.0.csv") %>% rename_all(tolower)

Part <- read_csv("~/Dropbox/projects/mid-project/internal/part-diagnostic-check.csv")

Conf %>%
  filter(fatalmin >= 1000) -> Wars

Conf %>%
  filter(fatalmin < 1000) -> notWars

TD <- read_csv("~/Dropbox/projects/mid-project/internal/td-diagnostic-check.csv") %>%
  left_join(., Conf %>% select(dispnum, is_war))
```

# Whither War? An Empirical Assessment of the Patterns of War
## Introduction
### Goal for Today

1. Discuss basic patterns in inter-state war.
2. Address the debate on the "decline of war."


### Confrontations and War

A militarized interstate confrontation (or dispute) is any threat, display, or use of force from one state to another.

- Gibler and Miller (Forthcoming): there have been 1,958 of these from 1816 to 2014.

Most of these (mercifully) don't go very far.

- The typical confrontation lasts a month or so and doesn't claim lives.
- Most common highest militarized actions: clash, attack, show of force.

### Confrontations and War

War has typically commanded the most attention in this field.

- Operationally: a subset of all confrontations with at least 1,000 battle-related deaths (excluding civilian deaths).

What does the typical war "look like?"

- If all you know about war is the world wars (or what the U.S. has done), you'd be misled about what war typically looks like.

Is war on the decline?

- Kinda depends on how you look at it (and *don't* lump all war types together).


## Basic Statistics About War
###

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  mutate(totalp = numa + numb) %>%
  summarize(#mode_mindur = mode(mindur),
            median_mindur = median(mindur),
            mean_mindur = mean(mindur),
            #mode_totalp = mode(totalp),
            median_totalp = median(totalp),
            mean_totalp = mean(totalp),
            #mode_fatalmin = mode(fatalmin),
            median_fatalmin = median(fatalmin),
            mean_fatalmin = mean(fatalmin),
            #mode_fatalax = mode(fatalmax),
            median_fatalmax = median(fatalmax),
            mean_fatalmax = mean(fatalmax)) -> war_summaries
```


| **Category** | **Median** | **Mean** | **Example** |
|:-------------|:----------:|:--------:|:------------|
| (Min.) Duration | `r war_summaries[1] %>% pull()` | `r round(war_summaries[2] %>% pull(), 3)` | First Kashmir War (438) | 
| Participants | `r war_summaries[3] %>% pull()` | `r round(war_summaries[4] %>% pull(), 3)` | First Taiwan Straits Crisis |
| (Min.) Fatalities | `r format(war_summaries[5] %>% pull(), big.mark=",", scientific=FALSE)` | `r format(round(war_summaries[6] %>% pull(), 3), big.mark=",", scientific=FALSE)` | Austro-Sardinian War (7304) |
| (Max.) Fatalities | `r format(war_summaries[7] %>% pull(), big.mark=",", scientific=FALSE)` | `r format(round(war_summaries[8] %>% pull(), 3), big.mark=",", scientific=FALSE)` |  Russian Civil War (12783) |

Table: Central Tendencies of Inter-state Wars (Gibler and Miller, Forthcoming)


###

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  mutate(durcat = case_when(
    mindur <= 366 ~ "1",
    between(mindur, 367, 732) ~ "2",
    between(mindur, 733, 1098) ~ "3",
    between(mindur, 1095, 1830) ~ "4-5",
    between(mindur, 1831, 3660) ~ "6-10",
    mindur >= 3661 ~ "10+"
  )) %>%
  group_by(durcat) %>%
  tally() %>%
  mutate(durcat = fct_relevel(durcat, "1", "2", "3", "4-5", "6-10", "10+")) %>%
  arrange(durcat) %>%
  ggplot(.,aes(durcat, n)) +
  geom_bar(stat="identity", color="black", alpha=0.8, fill="#619cff") +
  geom_text(aes(label=n), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  scale_x_discrete(labels = c("One Year\nor Less", "1-2 Years", "2-3 Years", "4-5\n(Full) Years", "6-10\n(Full) Years", "10 Years\nor More")) +
  theme_steve_web() +
  labs(title = "How Long do Inter-State Wars Last?",
       subtitle = "40% of wars last a year or less, and only about a third of inter-state wars last more than two years.",
       x = "", y = "",
       caption = "Data: Gibler and Miller (Forthcoming). Selects on all confrontations where minimum fatalities exceed 1,000.")
```

###


```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  mutate(totalp = numa + numb) %>%
  group_by(totalp) %>%
  tally() %>%
  ggplot(.,aes(as.factor(totalp), n)) +
  geom_bar(stat="identity", color="black", alpha=0.8, fill="#619cff") +
  geom_text(aes(label=n), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  theme_steve_web() +
  labs(title = "The Number of Participants in Inter-State Wars",
       subtitle = "Even here, the median misleads. The typical war is bilateral.",
       x = "Total Number of Participants", y = "",
       caption = "Data: Gibler and Miller (Forthcoming). Selects on all confrontations where minimum fatalities exceed 1,000.")
```


###

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  group_by(outcome) %>%
  tally() %>%
  #mutate(outcome = c("Victory for Side A", "Victory for Side B", "Yield by Side A", "Stalemate", "Compromise", "Joins Ongoing War"))
  ggplot(.,aes(as.factor(outcome), n)) +
  geom_bar(stat="identity", color="black", alpha=0.8, fill="#619cff") +
  geom_text(aes(label=n), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family="Open Sans") +
  scale_x_discrete(labels = c("Victory for\nSide A", "Victory for\nSide B", "Yield by\nSide A",
                              "Stalemate", "Compromise", "Joins Ongoing\nWar")) +
  theme_steve_web() +
  labs(title = "The Outcomes of Inter-State Wars",
       subtitle = "Wars typically end in a victory for one side, at least on the battlefield, but ~30% of wars end in a stalemate.",
       x = "", y = "",
       caption = "Data: Gibler and Miller (Forthcoming). Selects on all confrontations where minimum fatalities exceed 1,000.")

```


### Pop Quiz!

What do you think are:

- The deadliest wars (min/max fatalities)?
- The *least* deadly wars (min/max fatalities)?
- Longest/shortest wars?
- Most participants in wars?


### 

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  arrange(-fatalmax) %>%
  head(5) %>%
  select(dispnum, fatalmin, fatalmax) %>%
  mutate(warname = c("World War II", "World War I", "Second Sino-Japanese War",
                     "Vietnam War", "Korean War")) %>%
  select(dispnum, warname, everything()) %>%
  mutate(fatalmin = format(fatalmin, big.mark=",", scientific=FALSE),
         fatalmax = format(fatalmax, big.mark=",", scientific=FALSE)) %>%
  kbl(., booktabs = TRUE,
      align = c("c","l", "c", "c"),
      caption = "The Deadliest Inter-State Wars (Gibler and Miller, Forthcoming)",
      col.names = c("MIC No.", "War Name", "Min. Fatalities", "Max. Fatalities")) %>%
  row_spec(0, bold=TRUE)
```

*Bonus points to those who object the Paraguayan War should be in here.*

### 

```{r}
TD %>%
  filter(dispnum == 1590 & ccode1 == 150) %>%
  group_by(year) %>%
  summarize(fatalmin = sum(dyfatalmin1),
            fatalmax = sum(dyfatalmax1)) -> par_fatalities

par_fatalities %>%
  gather(var, val, -year) %>%
  group_by(year) %>%
  summarize(val = mean(val)/1000) %>%
  mutate(category = "Yearly Fatalities for Paraguay in Paraguayan War") -> par_fatalities

cow_nmc %>%
  filter(ccode == 150 & between(year, 1855, 1880)) %>%
  select(ccode, year, tpop) %>%
  mutate(category = "Total Population") %>%
  rename(val = tpop) %>%
  bind_rows(., par_fatalities) %>%
  ggplot(.,aes(as.factor(year), val, fill=category)) + 
  theme_steve_web() +
  geom_bar(stat="identity",color="black", position="dodge") +
  geom_text(aes(label=val), vjust=-.5, colour="black",
            position=position_dodge(.9), size=3.5, family="Open Sans") +
  labs(title = "The Paraguayan War Was Among the Biggest Single Disasters for Any State (Paraguay) in War",
       subtitle = "Paraguay lost around 30,000 troops a year in the seven-year war, lost about half its population, and disappeared from the state system for six years.",
       x = "Year", y = "People (in Thousands)",
       fill = "",
       caption = "Data: Gibler and Miller (Forthcoming), Correlates of War National Material Capabilities (v. 6.0)")

```

### 

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  arrange(fatalmin) %>%
  head(5) %>%
  select(dispnum, fatalmin, fatalmax) %>%
  mutate(warname = c("Second Taiwan Straits Crisis", "Football War", 
                     "The War of Attrition",
                     "Belgian Revolution", "Platine War")) %>%
  select(dispnum, warname, everything()) %>%
  mutate(fatalmin = format(fatalmin, big.mark=",", scientific=FALSE),
         fatalmax = format(fatalmax, big.mark=",", scientific=FALSE)) %>%
  kbl(., booktabs = TRUE,
      align = c("c","l", "c", "c"),
      caption = "The Least Deadly Inter-State Wars (Gibler and Miller, Forthcoming)",
      col.names = c("MIC No.", "War Name", "Min. Fatalities", "Max. Fatalities")) %>%
  row_spec(0, bold=TRUE)
```


### 

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  arrange(-maxdur) %>%
  head(5) %>%
  select(dispnum, mindur, maxdur) %>%
  mutate(warname = c('The Secret War', "Vietnam War", "Iran-Iraq War",
                     "Paraguayan War", "World War II")) %>%
  select(dispnum, warname, everything()) %>%
  mutate(mindur = format(mindur, big.mark=",", scientific=FALSE),
         maxdur = format(maxdur, big.mark=",", scientific=FALSE)) %>%
  kbl(., booktabs = TRUE,
      align = c("c","l", "c", "c"),
      caption = "The Longest Inter-State Wars (Gibler and Miller, Forthcoming)",
      col.names = c("MIC No.", "War Name", "Min. Duration", "Max. Duration")) %>%
  row_spec(0, bold=TRUE)
```


### 

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  arrange(mindur) %>%
  head(5) %>%
  select(dispnum, mindur, maxdur) %>%
  mutate(warname = c('Changkufeng Incident', "Hungarian Uprising", 
                     "Football War",
                     "Afghanistan War (!!)", 
                     "First Central American War")) %>%
  select(dispnum, warname, everything()) %>%
  mutate(mindur = format(mindur, big.mark=",", scientific=FALSE),
         maxdur = format(maxdur, big.mark=",", scientific=FALSE)) %>%
  kbl(., booktabs = TRUE,
      align = c("c","l", "c", "c"),
      caption = "The Shortest Inter-State Wars (Gibler and Miller, Forthcoming)",
      col.names = c("MIC No.", "War Name", "Min. Duration", "Max. Duration")) %>%
  row_spec(0, bold=TRUE)
```

The usual caveat about the Afghanistan War applies.

- Like the Iraq War, the *inter-*state component was brief. The *extra-*state component lasted forever.


### 

```{r}
Conf %>%
  filter(fatalmin >= 1000) %>%
  mutate(totalp = numa + numb) %>%
  arrange(-totalp) %>%
  head(5) %>%
  select(dispnum, numa, numb, totalp) %>%
  mutate(warname = c("World War II", "Iraq War", "Korean War",
                     "Afghanistan War", "World War I")) %>%
  select(dispnum, warname, everything()) %>%
  kbl(., booktabs = TRUE,
      align = c("c","l", "c", "c", "c"),
      caption = "The Most Participants in Inter-State Wars (Gibler and Miller, Forthcoming)",
      col.names = c("MIC No.", "War Name", "Num. Side A", "Num. Side B", "Total")) %>%
  row_spec(0, bold=TRUE)
```

## The Decline of War?
### The Decline of War?

Is war on the decline? Pinker (2011) says "yes."

- Argument: human beings are becoming less violent and more altruistic.
- The more pacific nature of post-WW2 is indicative of a powerful trend.

What can we say about this argument? And the overall trends of war?


### First Things First: Type Matters

It's a mistake to pool all war types together (c.f. Gibler and Miller, Forthcoming).

- *Inter-*: state(s) v. state(s)
- *Intra-*: state v. domestic non-state actor
- *Extra-*: state v. (in)dependent non-state actor
- *Non*-: non-state actor v. non-state actor


###

```{r}
cow_war_inter %>%
  group_by(warnum) %>%
  slice(1) %>%
  select(warnum, year) %>%
  mutate(cat = "Inter-state") %>%
  ungroup() -> inter

cow_war_intra %>%
  group_by(warnum) %>%
  slice(1) %>%
  select(warnum, year) %>%
  mutate(cat = "Intra-state") %>%
  ungroup() -> intra

cow_extra %>%
  select(warnum, startyear1) %>%
  group_by(warnum) %>%
  slice(1) %>%
  rename(year = startyear1) %>%
  mutate(cat = "Extra-state") %>%
  ungroup() -> extra

cow_non %>%
  select(warnum, startyear) %>%
  group_by(warnum) %>%
  slice(1) %>%
  rename(year = startyear) %>%
  mutate(cat = "Non-state") %>%
  ungroup() -> non

bind_rows(inter, intra, extra, non) %>%
  mutate(decade = paste0(year - year  %% 10, "s")) %>%
  group_by(cat, decade) %>%
  tally() %>%
  ggplot(.,aes(decade, n, fill=cat)) + 
  geom_bar(stat="identity",color="black", position="dodge") +
  geom_text(aes(label=n), vjust=-.5, colour="black",
            position=position_dodge(.9), size=3.5, family="Open Sans")  +
  theme_steve_web() +
  labs(title = "The Number of Unique War Onsets by Decade and CoW War Type, 1816-2007",
       subtitle = "The information here doesn't settle the decline-of-war debate, but it does caution against lumping all conflict forms together.",
       caption = "Data: Correlates of War War Data (v. 4.0 [v. 4.1 for intra-state wars])",
       x = "", y = "Number of Unique Onsets per Decade",
       fill = "")
```


### The Decline of War?

How do we quantify this?

- *magnitude*: (i.e. state-time in war)
- *severity*: raw battlefield deaths
- *intensity*: magnitude or severity, standardized
- *prevalence*: battle deaths, by world population


### 

```{r}
Conf %>%
  mutate(decade = paste0(styear - styear  %% 10, "s")) %>%
  filter(decade != "1810s" & decade != "2010s") %>%
  group_by(is_war, decade) %>%
  summarize(summindur = sum(mindur),
            summaxdur = sum(maxdur)) %>%
  ggplot(.,aes(decade, summindur, fill=is_war)) + 
  geom_bar(stat="identity",color="black", position= position_dodge2(preserve = "single")) +
  theme_steve_web() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "", y = "Total Duration (in Minimum Days)",
       fill = "",
       title = "Total Minimum Duration by Inter-State Confrontation Type and Decade of Confrontation Onset",
       subtitle = "The data don't suggest a clear declining pattern in terms of the magnitude of inter-state conflicts.",
       caption = "Data: Gibler and Miller (Forthcoming). Be mindful there are always going to be fewer wars than sub-war conflicts.")
```


### 

```{r}

TD %>%
  group_by(is_war, year) %>%
  summarize(sumfatalmin = sum(dyfatalmin1))  %>%
  ungroup() %>%
  #gather(var, val, -is_war, -year) %>%
  mutate(sumfatalmin = log(sumfatalmin + 1)) %>%
  ggplot(.,aes(year, sumfatalmin, linetype=is_war, color=is_war)) +
  geom_line(size=1.1) +
  facet_wrap(~is_war, nrow=2) +
  theme_steve_web() +
  labs(x = "", y = "Total Minimum Fatalities (Log-Transformed)",
       linetype = "", color="",
       title = "Total (Minimum, Log-Transformed) Fatalities by Inter-State Confrontation Type and Year",
       subtitle = "The data don't suggest a clear declining pattern in terms of the severity of inter-state conflicts.",
       caption = "Data: Gibler and Miller (Forthcoming). Be mindful there are always going to be fewer wars than sub-war conflicts.\nBraumoeller (2021) cautions against doing the quick-and-dirty transformation of variables drawn from a power law distribution.\nThat said, be mindful those two mountains in the middle of the war panel would be a lot larger if I didn't do this.")
```

### 

```{r}

TD %>%
  left_join(., Conf %>% select(dispnum, styear)) %>%
  select(dispnum, is_war, ccode1, year, styear, dyfatalmin1) %>%
  group_by(is_war, year) %>%
  summarize(sumfatalmin = sum(dyfatalmin1))  %>%
  ungroup() -> hold_this

cow_nmc %>%
  group_by(year) %>%
  summarize(sumtpop = sum(tpop)) %>%
  left_join(hold_this, .) %>%
  mutate(sumfatalmin = sumfatalmin/1000) %>%
  mutate(prop = log(sumfatalmin + 1)/log(sumtpop)) %>%
  ggplot(.,aes(year, prop, linetype=is_war, color=is_war)) +
  geom_line(size=1.1) +
  facet_wrap(~is_war, nrow=2, scales= "free") +
  theme_steve_web()  +
  labs(x = "", y = "Total Minimum Fatalities/World Population (Log-Transformed)",
       linetype = "", color="",
       title = "Total (Minimum, Log-Transformed) Fatalities over World Population by Inter-State Confrontation Type and Year",
       subtitle = "The data don't suggest a clear declining pattern in terms of the prevalence of inter-state conflicts.",
       caption = "Data: Gibler and Miller (Forthcoming) and Correlates of War National Material Capabalities (v. 6.0). Be mindful there are always going to be fewer wars than sub-war conflicts.\nBraumoeller (2021) cautions against doing the quick-and-dirty transformation of variables drawn from a power law distribution.\nThat said, be mindful those two mountains in the middle of the war panel would be a lot larger if I didn't do this.")

```


### Possible Objection

The decline-of-war advocates will typically note that we haven't seen a World War II in over 70 years.

- Which, sure, and thank goodness for that.
- There's also assuredly a peace that's grown from that (c.f. Gleditsch, 2002; Goertz et al., 2016).

Retort:

- We're talking about a rare event, and selecting on the rarest of those rare events.
- There's a problem in assessing central tendency in a distribution that is decidedly not normal.


### Another Way of Thinking About This

Gibler and Miller (Forthcoming) code 1,958 confrontations from 1816 to 2014. Of those:

- 84 are operationally a war (minimum fatalities >= 1000), 1,874 are not a war.
- The proportion of wars to all confrontations is .042.

Of the 84 wars:

- The median war claimed 7,478 troops in battle.
- The median absolute deviation is 8,367, the 95% interval of which maxes at 19,925.
- There are 24 wars with at least more than 19,925 fatalities (*p* = .012). 

Consistent with Pinker's (2011) argument, four of those are after World War II while the other 20 are before it.


### A Thought Experiment

The year is 2022. We haven't observed a World War II in, basically, 75 years. What is the probability of us *not* observing this where:

- *p* = .042, the overall base rate of war vs. not-war in the Gibler and Miller confrontation data?
- *p* = .012, the overall base rate of a "really big war" in the Gibler and Miller confrontation data?


###

```{r}
tibble(num_wars = seq(0:7)-1,
       base = dbinom(num_wars, 75, .042),
       rbw = dbinom(num_wars, 75, .012)) %>%
  gather(var, val, -num_wars) %>%
  mutate(var = ifelse(var == "base", "Base Rate of War", "Base Rate of 'Really Big Wars'")) %>%
  mutate(lab = round(val, 3)) %>%
  ggplot(.,aes(as.factor(num_wars), val, fill=var)) +
  geom_bar(stat = "identity", position = "dodge", color="black") +
  geom_text(aes(label=lab), vjust=-.5, colour="black",
            position=position_dodge(.9), size=3.5, family="Open Sans") +
  theme_steve_web() +
  labs(title = "The Probability of the Number of (Observed) Wars in 75 Years, Given Assumed Rates of War",
       subtitle = "Knowing how rare 'really big wars' are, it's highly probable (p = .404) that we haven't observed one 75 years after WW2.",
       fill = "",
       x = "Number of Observed Wars", y = "Probability of This Number of War in a 75-year Period")
```


###

```{r}

tibble(num_wars = rep(c(0, 1, 2), 100)) %>%
  arrange(num_wars) %>%
  mutate(period = rep(seq(1:100), 3),
         p = dbinom(num_wars, period, 0.012)) %>%
  mutate(cat = case_when(
    num_wars == 0 ~ "Zero Wars",
    num_wars == 1 ~ "One War",
    num_wars == 2 ~ "Two Wars"
  ),
  cat = fct_relevel(cat, "Zero Wars", "One War", "Two Wars")) %>%
  ggplot(.,aes(period, p, color=cat, linetype=cat)) +
  geom_line(size=1.1) +
  theme_steve_web() +
  labs(y = "Probability of Observing This Many Wars Over 100 Years",
       x = "",
       title = "The Probability of Observing a Set Amount of 'Really Big Wars' Over a 100-Year Period",
       color = "", linetype = "",
       subtitle = "After 75 years, it's still more probable that we haven't observed a 'really big war' than having observed just one.")

```

## Conclusion
### Conclusion

Most wars don't look like World War I and World War II, and thank goodness.

- The typical war is bilateral, about a year in length, with only a few thousand fatalities.
- That's not to say 'really big wars' aren't interesting. They're just not typical.

We still have a lot of work to do re: understanding systemic patterns in war.

- There's a clear post-WW2 peace, if in Western Europe (and Latin America to a lesser extent).
- There's no robust support for the "decline of war" hypothesis.
- *War varies by type, and **don't** lump them together*.

